---
title: "R intro"
output:
  pdf_document: default
  html_notebook: default
---

```{r}
getwd()
```

```{r}

```

```{r}
# will not print anything as its a comment

print("this will be printed")
```


### Arithmatics
```{r}
#Arithmatics. +,-,/, * are called operators
2 + 2
2 - 2
15 / 2
5 * 4
2 + 2 *3 # when using multiple operators calculations using BODMAS
(2 + 2) * 3
```

### R: VARIABLES
A variable is any text (letter/word/alpha-numeric) which stores some value
We use assignment operatr '<-' to pass a value to the variable
```{r}
# Variables
x <- 4
some_number <- 20
```



```{r}
# add two variables
x+some_number
```

```{r}
# change the value of a variable
x=100
x
x=x+2
x
```



### R: Vectors
Vectors are an collection of some data elements.
Vectors are also a form of variable but they contain a list of values. 
```{r}
 
some_vector1 <- c(1,2,3,4,5,6) # c is a 'function' that combines its arguments into a vector
some_vector1
some_vector2 <- 1:6 # a shortcut
some_vector2

series1 <- seq(from = 2, to = 20, by = 4) #series1 is a vector created by function seq
series1

```

```{r}
?seq
```

```{r}
#we can perform operations on vectors such as addition, multiplication etc
x <- 1:5
x
y= x*2
y
```

```{r}
# add the two vectors
joined_vectors <- some_vector1 + some_vector2 # the vectors will get added element-wise
joined_vectors
```

### indexing and manipulations

 - Square brackets [ ] - used for indexing (position)
 - Parentheses () - used in functions to provide arguements
```{r}

x=seq(1,12,2)
x


```
```{r}
x[3] # 3 here is the index value, i.e 3rd element in x
x[c(3,4)]
## removing an element
x[-3] ## is x changed?
y= x[-4]
y

which(x == 7)
```

### character vectors: Instead of numbers we use strings
```{r}
char_vec1 <- c("Archana", "Hanane", "Pallavi", "John")
char_vec1

```

### R: FUNCTIONS
Functions are a form of task performed on input 'arguments'
```{r}

## seq function is creating a series of numbers on the basis of specified parameters. 
x <- seq(from = 2, to = 20, by = 4)
x
length(x) # length is a function
```

### getting help

```{r}

?seq
```

# know the version
```{r}
#knowing the R version (plus the version of other attached packages)
sessionInfo() 
```
```{r}
# preview or knit the file in pdf or html document
```


